---
title: "9: Data Visualization"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse) #ggplot is included here

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.nutrients.gathered <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
PeterPaul.chem.nutrients.summaries <- read.csv("./Data/Processed/NTL-LTER_Lake_Summaries_PeterPaul_Processed.csv")
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1718_Processed.csv")

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d") #we always need to do the date thing again
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d") #we always need to do the date thing again Lubridate also can format dates.

```

## GGPLOT

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms #sometimes you can use other dataframes in the geom

Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

### Plotting continuous variables over time: Scatterplot
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()

O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot) # you get the same result. Print() prints it in a pdf.

# Fix this code
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone), color = "blue") # fixed
print(O3plot2)

# Add additional variables
PMplot <- 
  ggplot(EPAair, aes(x = as.factor(Month), y = PM25, shape = as.factor(Year), color = Site.Name)) +
  geom_point() # here makes sense color inside aes # month goes to 12.5. not pretty. as factor solves it 
print(PMplot)

# Separate plot with facets
PMplot.faceted <-
  ggplot(EPAair, aes(x = Month, y = PM25, shape = as.factor(Year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3) #new layer. three rows. three graphs. if not default is columns.
print(PMplot.faceted)

PMplot.faceted2 <-
  ggplot(EPAair, aes(x = Month, y = PM25)) +
  geom_point() +
  facet_grid(Site.Name ~ Year) # tilda is by
print(PMplot.faceted2)

# Filter dataset within plot building
O3plot.Blackstone <- 
  ggplot(subset(EPAair, Site.Name == "Blackstone"), aes(x = Date, y = Ozone)) + 
  geom_point() +
  geom_line() # puts a line between the dots
print(O3plot.Blackstone) #subset of only Blackstone

# Exercise: build your own scatterplots of PeterPaul.chem.nutrients

# 1. 
# Plot surface temperatures by day of  year. 
# Color your points by year, and facet by lake in two rows.

PeterPaul.chem.nutrients.faceted <-
  ggplot(PeterPaul.chem.nutrients, aes(x = daynum, y = temperature_C, color = year4)) +
  geom_point() +
  facet_wrap(vars(lakename), nrow = 2) 
print(PeterPaul.chem.nutrients.faceted)

#2. 
# Plot temperature by date. Color your points by depth.
# Change the size of your point to 0.5

PeterPaul.chem.nutrients.faceted2 <-
  ggplot(PeterPaul.chem.nutrients, aes(x = sampledate, y = temperature_C, color = depth)) +
  geom_point(size = 0.5)
  print(PeterPaul.chem.nutrients.faceted2)

```
### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

# Adjust axes
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) + # we set our axis to known data normal limits.
  ylim(0, 20)
print(lightvsDOfixed)

# Depth in the fields of limnology and oceanography is on a reverse scale
tempvsdepth <- 
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse()    #Reverse Axis
print(tempvsdepth)
#CLASS 2 ###########################################################################
NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +
  geom_point() +
  geom_smooth(method = lm) + #without the inside uses smooth function
  geom_abline(aes(slope = 16, intercept = 0), lty = 2) # lty = linetype. The points are above the line. We have more nitrogen than phosphorus. Typical in the US.
print(NvsP)   # there are grey points. points with no value for the color.

# Exercise: Plot relationships between air quality measurements

# 1. 
# Plot AQI values for ozone by PM2.5, colored by site. 

O3vsPM25 <-
  ggplot(EPAair, aes(x = PM25, y = Ozone, color = Site.Name)) +
  geom_point() #+
  geom_smooth(method = lm) #+ 
  geom_abline(aes(slope = 16, intercept = 0), lty = 2) 
print(O3vsPM25)


# Add a line of best fit for the linear regression of these variables.
O3vsPM25 <-
  ggplot(EPAair, aes(x = PM25, y = Ozone, color = Site.Name)) +
  geom_point() + # if you put here as(color=Site.Name) it gives one line for everything.
  geom_smooth(method = lm) #+ 
  geom_abline(aes(slope = 16, intercept = 0), lty = 2) 
print(O3vsPM25)

class(EPAair$Site.Name)
 
```

### Plotting continuous vs. categorical variables

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
 # Barplot + error bars
PeterPaul.nutrient.summaries <- PeterPaul.nutrients.gathered %>%
  group_by(lakename, nutrient) %>%   # we are grouping by two things
  summarise(sd = sd(concentration), 
            mean = mean(concentration))

Nutrientplot <- 
  ggplot(PeterPaul.nutrients.gathered) +
  geom_bar(aes(x = lakename, y = concentration, fill = nutrient), # why did we use fill? Instead of color. Color only paints the border of the bars.
           position = "dodge", stat = "summary", fun.y = "mean")  # what's happening here? Dodge = stuck on top of each other. stat --> Summary --> mean
print(Nutrientplot)

Nutrientplot2 <- 
  ggplot(PeterPaul.nutrient.summaries, aes(x = lakename, y = mean, fill = nutrient)) + #
  geom_bar(stat = "identity", position = "dodge") + # what does the stat command do? The indentity of the cell. We already calculated the mean. You get to the same bar plot.
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), # how do we specify error bars?
                 position = "dodge") #you need dodge
print(Nutrientplot2) # not useful. Error goes below to zero. Not real value.

# Are there more effective ways to produce summary stats for categories? Alternatives-->

# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"? Dont wanna fill only the box plot
print(Nutrientplot3)

# Dot plot
Nutrientplot4 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge") #
print(Nutrientplot4)

# Violin plot
Nutrientplot5 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "black") +   # if you dont put colors, all black
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "darkgray") +
  geom_freqpoly(aes(x = no23), color = "gray") +
  geom_freqpoly(aes(x = po4), color = "blue") 
print(Nutrientplot6)

# Using a gathered dataset    #gathered versus tidy dataset
Nutrientplot7 <-   
  ggplot(PeterPaul.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)

# Exercise: Plot distributions of AQI values for EPAair

# 1. 
# Create a bar chart plus standard deviation error bars for PM2.5, divided by year. 
# Create separate bars for each site. 

EPAairsummary <- EPAair %>%
  group_by(Site.Name, Year) %>%   #key for the summary
  summarise(mean=mean(PM25,na.rm = TRUE), 
            sd=sd(PM25,na.rm = TRUE))

AQIValuesEPAairbarplot <- 
  ggplot(EPAairsummary, aes(x = as.factor(Year), y = mean, fill = Site.Name)) + #
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd),
                 position = "dodge") 
print(AQIValuesEPAairbarplot) 

# 2. 
# Create a new plot that better depicts the distribution of PM2.5 concentrations. 
# Divide your graph by year and site.

```
